Characterizing the Splice Map of Turkey Hemorrhagic Enteritis Virus

### 2024-05-09 importance_of_study.Rmd
second round of edits after jvi review

\vspace{15mm}
Hemorrhagic enteritis (HE) is a disease of turkey poults characterized by immunosuppression, bloody diarrhea, and up to 80% mortality. This disease is caused by *Turkey Hemorrhagic Enteritis Virus* (THEV) of which avirulent strains that do not cause HE but retain the immunosuppressive ability have been isolated. The Virginia Avirulent Strain (VAS) is still used as a live vaccine despite its immunosuppressive properties. Thus, vaccinated birds are rendered more susceptible to opportunistic infections and death than unvaccinated cohorts. To establish the genetic basis by which VAS brings about immunosuppression leading to its mitigation, it is imperative that the viral gene(s) mediating the immunosuppression be well-characterized. As the viral splicing and gene expression patterns are unknown, the most pressing need was for a well-characterized transcriptome of THEV. Also, the detailed characterization of a non-human adenovirus splice map, which are scantily studied, provides valuable insights into the differences of various adenovirus splicing patterns.

>>>>

## List of abbreviations  
```{r echo=F, warning=F, message=F}
library(tidyverse)
library(flextable)
tribble(~Abbreviation, ~Definition,
               "AdV", "Adenovirus",
               "MAdV", "Mastadenovirus",
               "SiAdV", "Siadenovirus",
               "THEV", "Turkey Hemorrhagic Enteritis Virus",
              "HE", "Hemorrhagic Enteritis",
              "UTR", "Untranslated Region",
              "TU", "Transcription Unit",
              "L4P", "L4 Promoter",
              "MLP", "Major Late Promter",
              "E3P", "E3 Promoter",
              "hpi", "Hours Post-infection",
              "qPCR", "Quantitative Polymerase Chain Reaction",
              "FPKM", "Fragments Per Kilobase of transcript per Million mapped reads",
              "CP", "Coding Potential",
              "TSS", "Transcription Start Site",
              "TTS", "Transcription Termination Site",
              "SC", "Start Codon", 
              "STC", "Stop Codon",
              "secSC", "Secondary Start Codon",
              "secSTC", "Secondary Stop Codon",
              "ORF", "Open Reading Frame", 
              "CDS", "Coding Sequence", 
              "MLTU", "Major Late Transcription Unit",
              "TPL", "Tripartite Leader",
              "sTPL", "Short Tripartite Leader",
              "TPL3", "Third exon of Tripartite Leader",
              "GCN", "Genome Copy Number") %>%
  flextable::flextable() %>%
  flextable::theme_box() %>%
  flextable::width(width = 1, j = 1) %>%
  flextable::width(width = 2, j = 2) %>%
  flextable::align(align = "center", part = "all")
```

\newpage
Hemorrhagic enteritis (HE) is a disease of turkey poults characterized by immunosuppression, bloody diarrhea, and up to 80% mortality. This disease is caused by *Turkey Hemorrhagic Enteritis Virus* (THEV) of which avirulent strains that do not cause HE but retain the immunosuppressive ability have been isolated. The Virginia Avirulent Strain (VAS) is still used as a live vaccine despite its immunosuppressive properties. Thus, vaccinated birds are rendered more susceptible to opportunistic infections and death than unvaccinated cohorts. To establish the genetic basis by which VAS brings about immunosuppression leading to its mitigation, it is imperative that the viral gene(s) mediating the immunosuppression be well-characterized. As the viral splicing and gene expression patterns are unknown, the most pressing need was for a well-characterized transcriptome of THEV. Also, the detailed characterization of a non-human adenovirus splice map, which are scantily studied, provides valuable insights into the differences of various adenovirus splicing patterns.

### 2024-05-09 manuscript_thev_transcriptome.Rmd
second round of edits after jvi review

---
title: Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis Virus in a Turkey B-cell Line
output:
  officedown::rdocx_document:
    reference_docx: style_word_output.docx
  pdf_document:
    keep_tex: true
geometry: margin=1in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{placeins}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
  - \usepackage{caption}
  - \usepackage[labelformat=empty]{caption}
bibliography: transcriptome_refs.bib
csl: asm.csl
editor_options:
  chunck_output_type: console
  markdown: 
    wrap: sentence
---

@@

**Running Title:** Novel Insights into Turkey Hemorrhagic Enteritis Virus Transcriptome

\vspace{10mm}

Abraham Quaye^1\*^, Brett E. Pickett^\*^, Joel S. Griffitts^\*^, Bradford K. Berges^\*^, Brian D. Poole${^\dagger}$^\*^

@@

^\*^Department of Microbiology and Molecular Biology, Brigham Young University\
^1^First-author\
${^\dagger}$ Corresponding Author

@@

Characterizing the splice map of turkey hemorrhagic enteritis virus (THEV) is an essential step that would allow studies of individual genes mediating its immunosuppressive functions. We used an RNA-sequencing experiment to characterize the transcriptome of THEV for the first time, providing key insight into the THEV gene expression and mRNA structures. Researchers previously annotated the genome of THEV as encoding 23 open reading frames (ORFs). In this work we identified 29 spliced transcripts all of which consisted of novel exons although some exons matched some previously annotated ORFs. The three annotated splice junctions were also corroborated by our data. We performed PCR amplification of THEV cDNA, cloned the PCR products, and used Sanger sequencing to validate all identified splice junctions. During validation we identified five additional unique transcripts, a subset of which were further validated by 3' rapid amplification of cDNA ends (3' RACE) experiments. Thus, we report that the genome of THEV contains 34 transcripts with the coding capacity for all annotated ORFs. However, we found six of the previously annotated ORFs (ORF1, E3, 33K, ORF8, IVa2, and protease) to be truncated ORFs on the basis of the identification of an in-frame upstream start codon or the detection of additional coding exons. We also identified three of the annotated ORFs with longer or shorter isoforms, and seven novel unannotated ORFs that could potentially be translated; although it is beyond the scope of this manuscript to investigate whether they are translated. Similar to other adenoviruses (AdVs), THEV also produces multiple distinctly spliced transcripts that code for the same proteins across its genome. Our data show that all THEV transcripts are spliced and organized into five transcription units under the control of their cognate promoters like other AdVs.
\newpage

@@

A myriad of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs -- specifically human AdVs. Thus, most of the current AdV gene expression and replication knowledge is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV transcription is temporally regulated; their genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late transcription unit (MLTU or major late promoter [MLP] region), which generates five families of late mRNAs (L1-L5) based on the polyadenylation site. An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host-cell modulating proteins, primarily involved in DNA replication, or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins that act as capsid proteins, promote virion assembly, or direct genome packaging. The immediate early genes E1A are expressed first, followed by the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. It is noteworthy that the MLP shows basal transcriptional activity during early infection (before DNA replication), with a comparable efficiency to other early viral promoters, but it reaches its maximal activity during late infection (after DNA replication). However, during early infection only a subset of the MLP-derived transcripts are expressed [@Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. For instance, the MLTU produces over 20 distinct splice variants all containing three non-coding exons at the 5'-end (collectively known as the tripartite leader; TPL) [@Donovan2020; @Zhao2014]. There is also an alternate three-exon 5' non-coding leader sequence present in varying amounts on a subset of MLTU mRNAs (known as the x-, y-, and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5) based on polyadenylation site. Each transcription unit (TU) contains its own promoter driving the expression of the array of mRNA transcripts produced via alternative splicing in the unit [@Zhao2014; @Donovan2020; @Guimet2016]. The promoters are activated at different phases of the infection by proteins from previously activated TUs. Paradoxically, the early-to-late phase transition during infection requires the L4 gene products, 22K and 33K, which should only be available after the transition. However, a promoter in the L4 region (L4P) that directs the expression of these two proteins independent of the MLP was found, resolving the paradox [@Morris2010; @Guimet2016; @Donovan2020]. During translation of AdV mRNA, recent studies using long-read direct RNA sequencing strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020; @Westergren2021].

@@

**Early Region 2 (E2) and Intermediate Region (IM) transcripts**   
The E2 TU expressed on the anti-sense strand, is subdivided into E2A and E2B and encodes three classical AdV proteins -- pTP and Ad-pol (E2B proteins), and DBP (E2A protein) -- essential for genome replication [@Donovan2020; @Zhao2014]. Unlike MAdV where two promoters are known [@Donovan2020], we discovered only a single TSS (E2 TSS; 18,751 bp) for both E2A and E2B transcripts in THEV. However, similar to MAdVs, E2A and E2B transcripts have distinct TTSs, with E2B transcripts sharing the TTS of the IVa2 transcript of the IM region [@Donovan2020; @Zhao2014] (**Figure 7**).

@@

The E2B region transcripts also start with the E2-5'UTR but extend downstream to reach the TTS at 2334 bp in the IM region, which is in the immediate context of a canonical PAS (position 2333 bp) where polyadenylation likely occurs. The E2B transcripts, TRXPT_6 and TRXPT_7, are almost identical except for an extra splice junction at the 3'-end of TRXPT_6 (**Figure 7**). TRXPT_7 has the CP for both classical proteins (pTP and Ad-pol) encoded in this region, with the pTP ORF predicted to be spliced from two exons. The predicted splice junction of pTP is corroborated by our data but the full transcript is markedly longer than the predicted ORF, although, the encoded product (pTP) remains unchanged. Ad-pol (polymerase) is encoded downstream of pTP with secondary SC (secSC) usage. The CP of TRXPT_6 slightly differs from TRXPT_7 because a new STC resulting from the extra splice site forms a minimal truncation of the Ad-pol encoded from its secSC.

@@

**Early Region 3 (E3) transcripts**   
The E3 region, nested within the MLTU, encodes proteins that modulate and evade host immune defenses. In MAdVs, this region contains seven ORFs expressed from multiple transcripts sharing the same TSS (from the E3 promoter) but having different TTSs [@Guimet2016; @Donovan2020; @Zhao2014]. However, some E3 transcripts use the TSS of the MLP. Due to sharing the same TSS, in MAdVs, secSC usage is heavily relied on for gene expression in this region as utilizing only the first SC cannot produce the downstream transcripts in this TU. The 12.5K ORF and transcripts using the MLP TSS are exceptions [@Donovan2020].

@@

As the splice junctions of TRXPT_22, TRXPT_23, TRXPT_24, and TRXPT_29 share the same genomic space, their validation was done with a single primer pair, and they were differentiated from each other by cloning the cDNA and Sanger sequencing (**Supplementary PCR methods**). In addition to corroborating the splice junctions for the aforementioned transcripts, the Sanger sequencing results also showed a distinct splice variant undetected in our RNA-seq transcriptome. This was a three-exon transcript (TRXPT_30) with identical first and last exons as TRXPT_23, but also contained the second exon of TRXPT_24 (**Figure 8**). TRXPT_30 encodes a novel 140 residue, 15.7kDa protein (e22K), spanning all three exons. Interestingly, the last 81 C-terminus residues of e22K are identical to 22K (89 residues), a single-exon ORF predicted to use the same SC as 33K. Just as seen for 33K, the first 20 bp of 22K is intronic, excluding the first 7 residues of 22K from e22K. We consider e22K as a long variant of the predicted 22K ORF. Assuming TRXPT_30 shares the same TSS and TTS as TRXPT_23, it would also have the downstream CP of TRXPT_23.

@@

**Early Region 4 (E4) transcripts** 
This TU is found at the 3'-end of the genome and expressed on the anti-sense strand. Based on nucleotide position, ORF7 and ORF8 were predicted in this region [@Davison2003]; however, as ORF7 is neither on the anti-sense strand nor transcribed from a promoter in the E4 region; hence, we only classify ORF8 in this TU. This is corroborated by our RNA-seq data, showing only one transcript in this region on the anti-sense strand (**Figure 9**). The transcript (TRXPT_28) spans 25192-26247 bp and is spliced at 25701-26055 bp, forming a two-exon transcript. The second exon fully matches the predicted ORF8 with 12 extra base pairs at the 3'-end. However, we identified a SC 192 bp upstream of the predicted SC in the first exon from which an in-frame protein is encoded. We consider this longer isoform (eORF8 -- 26.4 kDa, 229 residues) as the genuinely expressed ORF. We also identified a canonical PAS 11 bp upstream of TTS (**Supplementary Table S2**).

@@

We identified five TTSs (10,549, 12,709, 16,870, 22,116, 25,168 bp) in this TU, which we consider as corresponding to the five late mRNA classes (L1-L5), respectively. L1 mRNAs include TRXPT_8, encoding the 52K ORF as predicted. L2 mRNAs include TRXPT_16, TRXPT_17, and TRXPT_18, all containing the sTPL with their respective coding exons. They encode pIIIa, III (penton), and pVII, respectively. The L3 mRNAs, TRXPT_14 and TRXPT_20, both encode the hexon (II) ORF but hexon is the only ORF encoded on TRXPT_14, whereas TRXPT_20 encodes pX (pre-Mu), pVI, and hexon in that order. L4 mRNAs, TRXPT_9, TRXPT_10, TRXPT_11, and TRXPT_13 are the largest transcripts in the transcriptome and encode several similar late proteins. TRXPT_9 and TRXPT_10 are very similar but not identical. The last exon of TRXPT_9 seems to be truncated and likely shares the same TTS as TRXPT_10. They both encode pVII as the 5'-most ORF and also have the CP for pX, pVI, hexon, a longer variant of protease (eProt) from an upstream in-frame SC, and ORF12 (a novel 120 residue protein). Additionally they have the CP for pVIII and eE3. TRXPT_11 encodes hexon as its 5'-most ORF and also has the CP for eProt, ORF12, e33K, pVIII and eE3. Typically, MLTU transcripts splice the TPL onto a splice site just upstream of the ORF to be expressed [@Donovan2020]. While this holds true for most MLTU ORFs, several late ORFs (pVI, protease, and ORF7) do not have such close proximity splicing but are contained in larger transcripts such as these L4 mRNAs, strongly suggesting the use of non-standard ribosomal initiation mechanisms such as secSC usage or ribosome shunting described in other AdVs for their translation [@Donovan2020; @Yueh1996]. TRXPT_13, an E3 ORF utilizing the MLP TSS, encodes the classical L4P genes, pVIII and eE3. Lastly, the L5 class transcript, TRXPT_12, encodes Fiber as its 5'-most ORF but also has the CP for ORF7. Interestingly, the CP of TRXPT_12 and TRXPT_27 of the E3 TU are identical but are initiated from different TSSs.
\newpage

## DISCUSSION/CONCLUSIONS

@@

Our work provides the first insights into THEV splicing, revealing 34 transcripts grouped into five transcription units (TUs). The general temporal gene expression regulation observed in MAdVs, with early regions peaking at earlier time points followed by MLTU predominance at later time points, seems to also apply to THEV. An unexpected observation is that the pileup of mapped reads to THEV seems consistently skewed over similar regions of the genome at all time points. Given the temporal regulation of AdVs gene expression, we anticipated distinct differences in read pileups over the genome at different time points, indicating the different stages of infection. This probably due to the infection not being well-synchronized, leading to transcripts overlapping the time points. Further research is required to determine the precise temporal regulation of THEV.

@@

AdV alternative splicing undergoes a regulated temporal shift in splice site usage, previously thought to be limited to certain TUs. However, recent studies suggest that AdVs routinely produce different combinations of splice acceptor–donor pairs across all TUs [@Westergren2021; @Akusjarvi2008; @Donovan2020; @Guimet2016]. The details of this phenomenon have been best studied for the E1A and L1 units. AdVs modulate the activities of the splicing factor U2AF and the cellular SR family of splicing factors (reviewed here [@Akusjarvi2008]), and encode several proteins that influence the RNA splice site used. This phenomenon appears to occur in the THEV transcriptome, as the stringency of splice acceptor-donor pairs selected decreases from the onset of the late phase (**Figure 5**). Recent studies show that a virtually unlimited number of combinatorial alternative splicing events occur in an AdV lytic infection, resulting in a variety of novel transcripts [@Donovan2020; @Westergren2021]. It is unlikely that the entire repertoire of mRNA produced via this mechanism will actually be translated. However, it has been speculated that the plasticity in alternative RNA splicing enables AdVs to fine-tune protein synthesis by providing different alternatively spliced variants encoding the same protein under changing conditions, confering an evolutionary advantage [@Donovan2020; @Westergren2021].

@@

## MATERIALS AND METHODS

>>>>

---
title: Characterizing the Splice Map of Turkey Hemorrhagic Enteritis Virus
output:
  officedown::rdocx_document:
    reference_docx: style_word_output.docx
  pdf_document:
    keep_tex: true
geometry: margin=1in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{placeins}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
  - \usepackage{caption}
  - \usepackage[labelformat=empty]{caption}
bibliography: transcriptome_refs.bib
csl: asm.csl
editor_options:
  chunck_output_type: console
  markdown: 
    wrap: sentence
---

@@

Abraham Quaye${^\dagger}$^,a^, Brett E. Pickett^a^, Joel S. Griffitts^a^, Bradford K. Berges^a^, Brian D. Poole^a,\*^

@@

^a^Department of Microbiology and Molecular Biology, Brigham Young University\
${^\dagger}$First-author\
^\*^Corresponding Author

@@

Characterizing the splice map of turkey hemorrhagic enteritis virus (THEV) is an essential step that would allow studies of individual genes mediating its immunosuppressive functions. We used RNA sequencing to characterize the transcriptome of THEV for the first time, providing key insights into the THEV gene expression and mRNA structures. Researchers previously annotated the genome of THEV as encoding 23 open reading frames (ORFs). In this work we identified 29 spliced transcripts all of which contained novel exons although some exons matched some previously annotated ORFs. The three annotated splice junctions were also corroborated by our data. We performed PCR amplification of THEV cDNA, cloned the PCR products, and used Sanger sequencing to validate all identified splice junctions. During validation we identified five additional unique transcripts, a subset of which were further validated by 3' rapid amplification of cDNA ends (3' RACE) experiments. Thus, we report that the genome of THEV contains 34 transcripts with the coding capacity for all annotated ORFs. However, we found six of the previously annotated ORFs (ORF1, E3, 33K, ORF8, IVa2, and protease) to be truncated ORFs on the basis of the identification of an in-frame upstream start codon or the detection of additional coding exons. We also identified three of the annotated ORFs with longer or shorter isoforms, and seven novel unannotated ORFs that could potentially be translated; although it is beyond the scope of this manuscript to investigate whether they are translated. Similar to other adenoviruses (AdVs), THEV also produces multiple distinctly spliced transcripts that code for the same protein. Our data show that all THEV transcripts are spliced and organized into five transcription units under the control of their cognate promoters like other AdVs.
\newpage

@@

A myriad of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs -- specifically human AdVs. Thus, most of the current AdV gene expression and replication knowledge is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV transcription is temporally regulated; their genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late transcription unit (MLTU or major late promoter [MLP] region), which generates five families of late mRNAs (L1-L5) based on the polyadenylation site. An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host-cell modulating proteins, primarily involved in DNA replication, or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins that act as capsid proteins, promote virion assembly, or direct genome packaging. The immediate early genes E1A are expressed first, followed by the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. It is noteworthy that the MLP shows basal transcriptional activity during early infection (before DNA replication), with a comparable efficiency to other early viral promoters, but it reaches its maximal activity during late infection (after DNA replication). However, during early infection only a subset of the MLP-derived transcripts are expressed [@Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. For instance, the MLTU produces over 20 distinct splice variants all containing three non-coding exons at the 5'-end (collectively known as the tripartite leader; TPL) [@Donovan2020; @Zhao2014]. There is also an alternate three-exon 5' non-coding leader sequence present in varying amounts on a subset of MLTU mRNAs (known as the x-, y-, and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5) based on polyadenylation site. Each transcription unit (TU) contains its own promoter driving the expression of the array of mRNA transcripts produced via alternative splicing in the unit [@Zhao2014; @Donovan2020; @Guimet2016]. The promoters are activated at different phases of infection by proteins from previously activated TUs. Paradoxically, the early-to-late phase transition during infection requires the L4 gene products, 22K and 33K, which should only be available after the transition. However, a promoter in the L4 region (L4P) that directs the expression of these two proteins independent of the MLP was found, resolving the paradox [@Morris2010; @Guimet2016; @Donovan2020]. During translation of AdV mRNA, recent studies using long-read direct RNA sequencing strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020; @Westergren2021].

@@

**Early Region 2 (E2) and Intermediate Region (IM) transcripts**   
The E2 TU expressed on the anti-sense strand, is subdivided into E2A and E2B and encodes three classical AdV proteins -- pTP and Ad-pol (E2B proteins), and DBP (E2A protein) -- essential for genome replication [@Donovan2020; @Zhao2014]. Unlike MAdV where two promoters are known [@Donovan2020], we discovered only a single TSS (E2 TSS; 18,751 bp) for both E2A and E2B transcripts in THEV. However, E2A and E2B transcripts have distinct TTSs, with E2B transcripts sharing the TTS of the IVa2 transcript of the IM region similar to MAdVs [@Donovan2020; @Zhao2014] (**Figure 7**).

@@

The E2B region transcripts also start with the E2-5'UTR but extend downstream to reach the TTS at 2334 bp in the IM region, which is in the immediate context of a canonical PAS (position 2333 bp) where polyadenylation likely occurs. The E2B transcripts, TRXPT_6 and TRXPT_7, are almost identical except for an extra splice junction at the 3'-end of TRXPT_6 (**Figure 7**). TRXPT_7 has the CP for both classical proteins (pTP and Ad-pol) encoded in this region, with the pTP ORF predicted to be spliced from two exons. The predicted splice junction of pTP is corroborated by our data but the full transcript is markedly longer than the predicted ORF, although the encoded product (pTP) remains unchanged. Ad-pol (polymerase) is encoded downstream of pTP with secondary start codon (secSC) usage. The CP of TRXPT_6 slightly differs from TRXPT_7 because a new STC resulting from the extra splice site forms a minimal truncation of the Ad-pol encoded from its secSC.

@@

**Early Region 3 (E3) transcripts**   
The E3 region, nested within the MLTU, encodes proteins that modulate and evade host immune defenses. In MAdVs, this region contains seven ORFs expressed from multiple transcripts sharing the same TSS (from the E3 promoter) but having different TTSs [@Guimet2016; @Donovan2020; @Zhao2014]. However, some E3 transcripts use the TSS of the MLP. Due to sharing the same TSS, in MAdVs, secSC usage is heavily relied on for gene expression in this region as utilizing only the first SC cannot produce the downstream proteins in this TU. The 12.5K ORF and transcripts using the MLP TSS are exceptions [@Donovan2020].

@@

As the splice junctions of TRXPT_22, TRXPT_23, TRXPT_24, and TRXPT_29 share the same genomic space, their validation was done with a single primer pair, and they were differentiated from each other by cloning the cDNA and Sanger sequencing (**Supplementary PCR methods**). In addition to corroborating the splice junctions for the aforementioned transcripts, the Sanger sequencing results also showed a distinct splice variant undetected in our RNA-seq transcriptome. This was a three-exon transcript (TRXPT_30) with identical first and last exons as TRXPT_23, which also contained the second exon of TRXPT_24 (**Figure 8**). TRXPT_30 encodes a novel 140 residue, 15.7kDa protein (e22K), spanning all three exons. Interestingly, the last 81 C-terminus residues of e22K are identical to 22K (89 residues), a single-exon ORF predicted to use the same SC as 33K. Just as seen for 33K, the first 20 bp of 22K is intronic, excluding the first 7 residues of 22K from e22K. We consider e22K as a long variant of the predicted 22K ORF. Assuming TRXPT_30 shares the same TSS and TTS as TRXPT_23, it would also have the downstream CP of TRXPT_23.

@@

**Early Region 4 (E4) transcripts** 
This TU is found at the 3'-end of the genome and expressed on the anti-sense strand. Based on nucleotide position, ORF7 and ORF8 were predicted in this region [@Davison2003]; however, as ORF7 is neither on the anti-sense strand nor transcribed from a promoter in the E4 region, we only classify ORF8 in this TU. This is corroborated by our RNA-seq data, showing only one transcript in this region on the anti-sense strand (**Figure 9**). The transcript (TRXPT_28) spans 25192-26247 bp and is spliced at 25701-26055 bp, forming a two-exon transcript. The second exon fully matches the predicted ORF8 with 12 extra base pairs at the 3'-end. However, we identified a SC 192 bp upstream of the predicted SC in the first exon from which an in-frame protein is encoded. We consider this longer isoform (eORF8 -- 26.4 kDa, 229 residues) as the genuinely expressed ORF. We also identified a canonical PAS 11 bp upstream of TTS (**Supplementary Table S2**).

@@

We identified five TTSs (10,549, 12,709, 16,870, 22,116, 25,168 bp) in this TU, which we consider as corresponding to the five late mRNA classes (L1-L5), respectively. L1 mRNAs include TRXPT_8, encoding the 52K ORF as predicted. L2 mRNAs include TRXPT_16, TRXPT_17, and TRXPT_18, all containing the sTPL with their respective coding exons. They encode pIIIa, III (penton), and pVII, respectively. The L3 mRNAs, TRXPT_14 and TRXPT_20, both encode the hexon (II) ORF but hexon is the only ORF encoded on TRXPT_14, whereas TRXPT_20 encodes pX (pre-Mu), pVI, and hexon in that order. L4 mRNAs, TRXPT_9, TRXPT_10, TRXPT_11, and TRXPT_13 are the largest transcripts in the transcriptome and encode several similar late proteins. TRXPT_9 and TRXPT_10 are very similar but not identical. The last exon of TRXPT_9 seems to be truncated and likely shares the same TTS as TRXPT_10. They both encode pVII as the 5'-most ORF and also have the CP for pX, pVI, hexon, a longer variant of protease (eProt) from an upstream in-frame SC, and ORF12 (a novel 120 residue protein). Additionally, they have the CP for pVIII and eE3. TRXPT_11 encodes hexon as its 5'-most ORF and also has the CP for eProt, ORF12, e33K, pVIII and eE3. Typically, MLTU transcripts splice the TPL onto a splice site just upstream of the ORF to be expressed [@Donovan2020]. While this holds true for most MLTU ORFs, several late ORFs (pVI, protease, and ORF7) do not have such close proximity splicing but are contained in larger transcripts such as these L4 mRNAs, strongly suggesting the use of non-standard ribosomal initiation mechanisms such as secSC usage or ribosome shunting described in other AdVs for their translation [@Donovan2020; @Yueh1996]. TRXPT_13, an E3 ORF utilizing the MLP TSS, encodes the classical L4P genes, pVIII and eE3. Lastly, the L5 class transcript, TRXPT_12, encodes Fiber as its 5'-most ORF but also has the CP for ORF7. Interestingly, the CP of TRXPT_12 and TRXPT_27 of the E3 TU are identical but are initiated from different TSSs.
\newpage

## DISCUSSION

@@

Our work provides the first insights into THEV splicing, revealing 34 transcripts grouped into five transcription units (TUs). The general temporal gene expression regulation observed in MAdVs, with early regions peaking at earlier time points followed by MLTU predominance at later time points, seems to also apply to THEV. An unexpected observation is that the pileup of mapped reads to THEV seems consistently skewed over similar regions of the genome at all time points. Given the temporal regulation of AdVs gene expression, we anticipated distinct differences in read pileups over the genome at different time points, indicating the different stages of infection. This may be due to an unsynchronized infection, leading to transcripts overlapping the time points. Further research is required to determine the precise temporal regulation of THEV.

@@

AdV alternative splicing undergoes a regulated temporal shift in splice site usage, previously thought to be limited to certain TUs. However, recent studies suggest that AdVs routinely produce different combinations of splice acceptor–donor pairs across all TUs [@Westergren2021; @Akusjarvi2008; @Donovan2020; @Guimet2016]. The details of this phenomenon have been best studied for the E1A and L1 units. AdVs modulate the activities of the splicing factor U2AF and the cellular SR family of splicing factors (reviewed here [@Akusjarvi2008]), and encode several proteins that influence the RNA splice site used. This phenomenon appears to occur in the THEV transcriptome, as the stringency of splice acceptor-donor pairs selected decreases from the onset of the late phase (**Figure 5**). Recent studies show that a virtually unlimited number of combinatorial alternative splicing events occur in an AdV lytic infection, resulting in a variety of novel transcripts [@Donovan2020; @Westergren2021]. It is unlikely that the entire repertoire of mRNA produced via this mechanism will actually be translated. However, it has been speculated that the plasticity in alternative RNA splicing enables AdVs to fine-tune protein synthesis by providing different alternatively spliced variants encoding the same protein under changing conditions, conferring an evolutionary advantage [@Donovan2020; @Westergren2021].

@@

## METHODS
